---
title: "Exponential Distribution & Central Limit Theorem"
author: "John James"
date: "March 25, 2016"
output:
  html_document:
    fig_caption: yes
    keep_md: yes
    theme: journal
  word_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kfigr)
library(printr)
library(reshape2)
read_chunk("exponentials.R")
knitr::opts_chunk$set(fig.path = "../figures/", fig.height = 4, fig.width = 10, echo=TRUE, warning=FALSE)
```


```{r environment, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<environment>>
```

```{r theoreticalValues, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<theoreticalValues>>
```

```{r sampleMeans, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<sampleMeans>>
```

```{r compare, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<compare>>
```

```{r t.test, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<t.test>>
```


```{r sample1000, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
<<sample1000>>
```

#Introduction
The exponential distribution is the probability distribution, a specific case of the gamma distribution, that describes the time between events in a Poisson process, i.e. a process in which events occur continuously and independently at a constant average rate.[1]  

The central limit theorem (CLT) states that, given certain conditions, the arithmetic mean of a sufficiently large sample size, n, of independent random variables, each with a well-defined expected value and well-defined variance, will be approximately normally distributed, regardless of the underlying distribution.[2] Moreover, the mean of a sampling distribution of means will appoximate the population mean and the sampling distribution variance will approximate the population variance divided by the sample size.

This paper examines properties of the exponential distribution vis-a-vis the central limit theorem. Specifically, the paper will address, via simulation and the application of the CLT, the following **research questions:**

1. What is the relationship between **µ**, the **theoretical mean (1/λ)** of the exponential distribution, and the **sampling distribution of means** of randomly generated exponentials?
2. How variable is the **sampling distribution of means** vis-a-vis the **theoretical mean** with expected variance equal to **(1/n)(1/(λ^2^)**?
3. Is the sampling distribution of means approximately normal?  More specifically, does the sampling distribution approximate a normal distribution with mean **µ** and variance **σ^2^/n**? 
4. What is the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of exponentials?

#Methods
To create a sampling distribution of means of n = 40 randomly generated exponentials, 4000 random exponents were generated, using the the rexp function in R, and stored in a 1000 x 40 matrix. Using the apply function, the mean of each row of 40 exponentials was calculated creating a **sampling distribution of means** of size N = 1000 (`r figr("sampleMeansCode", TRUE, type="Code")`). To ascertain the **equivalency of the theoretical mean and the sampling distribution mean**, A one sample t-test was conducted (`r figr("t.testCode", TRUE, type="Code")`) and a graphical representation of the sampling distribution vis-a-vis the theoretical mean was rendered. To evaluate **variability of the sampling and theoretical distributions**, the mean, variance and standard error of the sampling distribution were calculated and presented vis-a-vis the theoretical expected values and densities of the theoretical and sampling distributions were prepared to graphically compare the relative distribution variances. A normal quartile plot was rendered to show **normality of the sampling distribution**. Lastly, to illuminate the difference between the distribution of a **large collection of random exponentials** and the distribution of a **large collection of averages** of 40 exponentials, a large collection of n = 1000 exponentials was generated (`r figr("sample1000", TRUE, type="Code")`) and comparative density plots were prepared and rendered. 

#Analysis

##Sample Mean Versus Theoretical Mean
According to the central limit theorem, the mean of the sampling distribution (sample mean) should approximate the theoretical mean of the distribution.  The theoretical mean for an exponential distribution with λ = 0.2 is 5.  The following hypothesis was tested to show equivalency.

    H0: X =  μ  Where {X is the sample mean} 
    Ha: X <> μ        {μ is the theoretical mean for λ = 0.2}
    
The table below shows the sample mean to be within the 95% confidence interval of the theoretical mean.  Further, the p.value is significantly higher than α = 0.5.  Therefore the null hypothesis is not rejected.  The sample mean is statistically equivalent to the theoretical mean.  

```{r meanTest, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
meanTest
```

`r figr("densityPlot", TRUE, type="Figure")` shows the sampling distribution of means about the theoretical mean, depicted by the blue vertical line. 

##Sample Variance versus Theoretical Variance
The table below shoes that the equivalency of the variances of the sample and theoretical distributions. 
```{r comparetbl, echo = FALSE, message = F, eval = T, tidy=TRUE, tidy.opts=list(comment=TRUE)}
compare
```

`r figr("dualDensities1", TRUE, type="Figure")` juxtaposes the sampling and theoretical distributions of the means, graphically demonstrating the equality of the variances of the distributions. 

##Distribution of Sample Means
`r figr("nqp", TRUE, type="Figure")` shows a normal quartile plot for the sampling distribution. As predicted by the central limit theorem, the sampling distribution of means is indeed normal.

##Distribution of Exponentials versus Distribution of Means
The purpose of this section is to illuminate the difference between the distribution of a **large collection of random exponentials** and the distribution of a **large collection of averages** of 40 exponentials.  The distribution of a large collection of random exponentials assumes a gamma distribution, whereas the distribution of means assumes a normal distribution concentrated on the theoretical mean as graphically presented in `r figr("dualDensities2", TRUE, type="Figure")`.  Moreover, the variance of the sampling distribution is (1/n) times the variance of the distribution of exponents.

#Conclusion
The following summarizes the answers to the research questions presented.

1. What is the relationship between **µ**, the **theoretical mean (1/λ)** of the exponential distribution and the **sampling distribution of means** of randomly generated, exponentials?
    + The sample mean approximates the theoretical mean µ; moreover, the sampling distribution of means is normal about the theoretical mean.  
  
2. How variable is the **sampling distribution of means** vis-a-vis the **theoretical mean** with expected variance equal to **(1/n)(1/(λ^2^)**?
    + The variability of the sample mean closely approximates the *expected* variability of the theoretical variance.
  
3. Is the distribution of the sample mean approximately normal?  More specifically, does the distribution of the sample mean approximate a normal distribution with mean **µ** and variance **σ^2^/n**? 
    + The sampling distribution of means is normal about the theoretical mean; moreover, the distribution approximates a normal distribution with mean **µ** and variance **σ^2^/n**.

4. What is the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of exponentials?
    + Both distributions estimate the mean; however, the collection of means is normally distributed and has a variance (1/n) times the variance of the distribution of exponents, demonstrating that more sampling data results in less variability in the data. 


##References
1. Wikipedia. (2016, March 25). Exponential Distribution. Retrieved from Wikipedia.org: https://en.wikipedia.org/wiki/Exponential_distribution
2. Wikipedia. (2016, March 25). Central Limit Theorem (CLT). Retrieved from Wikipedia:     https://en.wikipedia.org/wiki/Central_limit_theorem#Classical_CLT

#Appendix

##Figures
```{r densityPlot, anchor = "Figure", echo = FALSE, message = F, eval = T}
<<densityPlot>>
```

`r figr("densityPlot", TRUE, type="Figure")`: Density of Samples Means 

```{r dualDensities1, anchor = "Figure", echo = FALSE, message = F, eval = T}
<<dualDensities>>
```

`r figr("dualDensities1", TRUE, type="Figure")`: Density of Theoretical versus Sample Distributions

```{r nqp, anchor = "Figure", echo = FALSE, message = F, eval = T, tidy=TRUE}
<<nqp>>
```

`r figr("nqp", TRUE, type="Figure")`: Normal Quartile Plot of Empirical Means (n=1000)

```{r dualDensities2, anchor = "Figure", echo = FALSE, message = F, eval = T}
<<dualDensities2>>
```

`r figr("dualDensities2", TRUE, type="Figure")`: Sampling Distribution vis-a-vis Distribution of Exponentials


##Code
`r figr("sampleMeansCode", TRUE, type="Code")`: Code to simulate and create sample distribution of means

```{r sampleMeansCode, anchor = "Code", echo = TRUE, message = F, eval = F}
<<sampleMeans>>
```

`r figr("t.testCode", TRUE, type="Code")`: Code to conduct one-sample t.test of sample mean vis-a-vis theoretical mean

```{r t.testCode, anchor = "Code", echo = TRUE, message = F, eval = F}
<<t.test>>
```

`r figr("sample1000Code", TRUE, type="Code")`: Code to render normal quartile plot of sample distribution of means

```{r sample1000Code, anchor = "Code", echo = TRUE, message = F, eval = F}
<<sample1000>>
```
